prolfqua (prodea) - An R-package for Proteomics Differential Expression Analysis.

### 2022-04-05 inst/MyArticle/paper.Rmd
added table, adressed TODO's

---
title: prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments
author: 
  - Witold E. Wolski^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich]
  - Paolo Nanni$^*$
  - Jonas Grossmann$^*$
  - Maria d'Errico$^*$
  - Ralph Schlapbach$^*$
  - Christian Panse$^*$
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
  Protein mass spectrometry is widely used for quantitative proteomics
  studies and in particular for relative protein quantification.
  Nevertheless, there is still a need for a flexible and easy-to-use, 
  transparently supporting a variety of well principled statistical
  inference procedures data analysis application programmer interface
  in R. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates
  essential steps of the mass spectrometry-based label-free quantification
  data analysis workflow as: quality control, data normalization, protein
  aggregation, statistical modeling, hypothesis testing, and sample size
  estimation.
  We use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to visualize and model
  simple experimental designs with a single explanatory variable and complex
  experiments with multiple factors.
  Furthermore, we use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to benchmark
  data acquisition, data preprocessing or data modeling methods. Furthermore,
  we apply `r BiocStyle::Githubpkg("fgcz/prolfqua")` to develop highly
  customizable, visually appealing, and interactive data analysis reports
  in PDF or HTML format.
  The R package is available through https://github.com/fgcz/prolfqua.
---

@@

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. We obtain different fold changes by comparing the various dilutions, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected fold change is 1.2 for E. coli proteins and 1 for human proteins.
We processed the raw data using MaxQuant [@Cox2008MaxQuant] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by Maxquant are available in the _prolfquadata_ R package [@prolfquadata].

@@

@@

# TODO(wew): give tmp a meaningful name
tmp <- prolfqua::addContrastResults(prefer = contr, add = conI)

## visualize results
pl <- tmp$merged$get_Plotter()
p1 <- pl$histogram()$p.value
p2 <- pl$volcano()$FDR +
   ggplot2::theme(legend.position = "bottom")

@@

We can use various statistics to examine the alternative hypothesis $\beta \ne 0$: the contrast estimate, i.e. the $\log_2$ fold-change $\beta$, the $t$-statistic $\frac{\beta}{\sqrt{var(\beta)}}$, or the $p$-value and moderated $p$-value. For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)). From the confusion matrix we obtain measures such as true positive rate ($TPR$), false positive rate ($FPR$), or false discovery proportion ($FDP$) which are given by:

@@

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest AUC scores among all the tested models. By using the beta distribution to model the number of peptides observed, we can further improve the pAUC scores (see Figure TODO(wew)). However, the properties of beta-based probabilities are not well understood; for instance, they have no uniform distribution under the null hypothesis (TODO remove they). The FDR estimates obtained when correcting multiple testing with the Benjamini-Hochberg method are strongly biased (see Figure \@ref(fig:FDRfdp) right panel). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. Furthermore, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset (see Figure \@ref(fig:FDRfdp) left panel). To address the issue of missing observation, we could either run `r BiocStyle::Biocpkg("proDA")` or the imputation of group averages method discussed here and afterwards apply `r BiocStyle::Biocpkg("PECA")`.

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available (for an overview see table TODO(wew) create table with methods implemented in prolfqua), and the contrast specification is consistent and intuitive for all models. The benchmarking results are comparable with existing methods. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for customizing the analysis or adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), quickly.

@@

TODO(wew)

>>>>

---
title: prolfqua (prodea) - An R-package for Proteomics Differential Expression Analysis.
author: 
  - Witold E. Wolski^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - Swiss Federal Institute of Technology in Zurich \| University of Zurich]
  - Paolo Nanni$^*$
  - Jonas Grossmann$^*$
  - Maria d'Errico$^*$
  - Ralph Schlapbach$^*$
  - Christian Panse$^*$
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
  Protein mass spectrometry is widely used for quantitative proteomics
  studies and in particular for relative protein quantification.
  Nevertheless, there is still a need for a flexible and easy-to-use, 
  transparently supporting a variety of well principled statistical
  inference procedures data analysis application programmer interface
  in R. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates
  essential steps of the mass spectrometry-based label-free quantification
  data analysis workflow as: quality control, data normalization, protein
  aggregation, statistical modeling, hypothesis testing, and sample size
  estimation.
  We use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to visualize and model
  simple experimental designs with a single explanatory variable and complex
  experiments with multiple factors.
  Furthermore, we use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to benchmark
  data acquisition, data preprocessing or data modeling methods. Furthermore,
  we apply `r BiocStyle::Githubpkg("fgcz/prolfqua")` to develop highly
  customizable, visually appealing, and interactive data analysis reports
  in PDF or HTML format.
  The R package is available through https://github.com/fgcz/prolfqua.
---

@@

To evaluate the performance of differential expression analysis, we use the _IonStar_ benchmark dataset[@shen2018ionstar], available from the Proteomics Identifications Database (PRIDE) identifier PXD003881. $DH5\alpha$ _E. coli_ lysate was spiked in human pancreatic cancer cells(Panc-1) lysate at five different levels: $3\%$ _E. coli_, $4.5\%$ _E. coli_, $6\%$ _E. coli_, $7.5\%$ _E. coli_ and $9\%$ _E. coli_. We annotated these dilutions from smallest to largest with the letters a, b, c, d, e. As a result, we obtain different fold changes by comparing the various dilutions, e.g., when comparing dilution _e_ (9%) against dilution _d_ (7.5%), the expected fold change is 1.2 for E. coli proteins and 1 for human proteins.
We processed the raw data using MaxQuant [cite] version Version 1.6.10.43, with MaxQuant default settings. The text files generated by Maxquant are available in the _prolfquadata_ R package [@prolfquadata].

@@

```{r prolfquaModels}
xM <- data.frame( 
    lm = c("lm", "linear model on peptide or protein abundances", "strategy_lm, build_models, Contrasts"),
    lmer = c("lmer" , "mixed effect models on peptide and protein abundances", "strategy_lmer, build,models, Contrasts"),
    gmi = c("group mean imputation", "impute missing group means", "ContrastsSimpleImpute"),
    ROPECA = c("ROPECA","linear models on peptide abundances", "strategy_lm, build_models, Contrast, ContrastsROPECA"),
    limma = c("limma" , "empirical Bayes variance shrinkage", "ContrastsModerated"),
    pic = c("Protein interaction scoring", "by wrapping SAINT express binaries", "runSaint, ContrastsSaintExpress"),
    prodA = c("dropout model", "by warpping proDA functionality", "strategy_proDA, build_models, ContrastsProDa"), check.names =  FALSE)
xM <- data.frame(t(xM))
colnames(xM) <- c("m", "model", "prolfqua functions")
knitr::kable(xM[,2:3], caption = "Models available in prolfqua.")

@@

contrasts <- prolfqua::addContrastResults(prefer = contr, add = conI)

## visualize results
pl <- contrasts$merged$get_Plotter()
p1 <- pl$histogram()$p.value
p2 <- pl$volcano()$FDR +
   ggplot2::theme(legend.position = "bottom")

@@

We can use various statistics to examine the alternative hypothesis $\beta \ne 0$: the contrast estimate, i.e. the $\log_2$ fold-change $\beta$, the $t$-statistic $\frac{\beta}{\sqrt{var(\beta)}}$, or the $p$-value and moderated $p$-value. For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:knitrConfusionMatrix)). From the confusion matrix we obtain measures such as true positive rate ($TPR$), false positive rate ($FPR$), or false discovery proportion ($FDP$) which are given by:

@@

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest AUC scores among all the tested models. By using the beta distribution to model the number of peptides observed, we can further improve the pAUC scores (see Figure \@ref(fig:FDRfdp) top). However, the properties of beta-based probabilities are not well understood; for instance, the p-values are not uniformely distributed under the null hypothesis (not shown). Furthermore, the FDR estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased overestimating the false discovery proprotion (see Figure \@ref(fig:FDRfdp) right panel). Therefore, we can not recommend this method if an unbiased estimate of FDR is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset (see Figure \@ref(fig:FDRfdp) bottom left). To address the issue of missing observation, we could either run `r BiocStyle::Biocpkg("proDA")` or the imputation of group averages method discussed here and afterwards apply `r BiocStyle::Biocpkg("PECA")`.

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is consistent and intuitive for all models. The benchmarking results are comparable with existing methods. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for customizing the analysis or adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), quickly.

@@

```{r}
